Classifying Proper Exercise From Sensory ata
========================================================

## Synopsis
Correct form is critical for reducing the risk of injury from exercises that use weights. An injury can negatively impact further training and led to less physical fitness and well-being. Using sensorory data provided from a Microsoft Xbox connect we can classify whether a subject is correctly performing an exercise and classify the specific mistake if the repitition is not correct.

```{r echo=TRUE}
library(ggplot2)
suppressMessages(library(caret))
library(ggplot2); library(caret)
# set R so that scientific notation less likely
options(scipen = 5)
```
```{r echo=FALSE}
### set work directory based on the machine I am using
if (file.exists("F:/machine_learning/assignments/ml_project"))  {
  setwd("f:/machine_learning/assignments/ml_project")
} else if (file.exists("E:/machine_learning/assignments/ml_project")){
  setwd("e:/machine_learning/assignments/ml_project")
} else {
  stop("no directory")
}
```


## Data Processing
### Read in the data from the original bz2 file
```{r echo=TRUE,cache=TRUE}
### Data Manipulation
# read the data
pml <- read.csv("./raw_data/pml-training.csv",stringsAsFactors=FALSE)
```

```{r echo=TRUE}
# eliminate variables with low variability 
# using the nearZeroVar function of the caret package
nz <- nearZeroVar(pml)
pml <- pml[,-nz]

# get rid of columns that are mostly NA
pml <- pml[,colSums(is.na(pml))/nrow(pml)<.9]

# get rid of user name and time related variables
pml <- pml[,-c(1:6)]

# make classe a factor variable
pml$classe <- factor(pml$classe)

# split data into training and test sets
set.seed(77)
inTrain <- createDataPartition(y=pml$classe,p=0.6, list=FALSE)
training <- pml[inTrain,]
testing <- pml[-inTrain,]
```

## Results
### Summarize and reshape data for reporting and chart creation
```{r }
```

## Summary
```{r tornado_location_plot,fig.width=8, fig.height=6}
                                ,data = df) + xlab("") + ylab("")
```


# References
1. Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.
2. Kuhn, Max Building Predictive Models in R Using the caret Package. Journal of Statistical Software. Vol. 28. November 2008.
3. Breiman, L; Cutler, A; Random Forests. http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm